# Brushing Up on the Basics

## Mixed ANOVA *R* {#mixed-anova-r}

Below I have conducted a mixed ANOVA based on the data from the dataset `iris`.

The **dependent variable** used in said calculation are the **respective measurements** recorded in the dataset.

**"Species"** is the **between-factor**, with "Setosa", "Versicolor" and "Virginica" being its factor levels.

**"Part of Flower"** is the **first within-factor** with "Petal" and "Sepal" being its factor levels.

**"Kind of Measurement"** is the **second within-factor** with "Width" and "Length" being its factor levels.

### Preperation of Dataset `iris` *R* {#preperation-of-dataset-iris-r}

Before a mixed ANOVA analysis can be conducted the data in the dataset `iris` needs to be reordered and recoded accordingly.

```{r}
#| label: anova-reoreco
#| tidy: true
#| tidy-opts:
#|   - width.cutoff=60
#| warning: false
#| message: false

# Create custom function to recode variables 
 recode.var <- function(x,   # what vector do you want to recode?
                       old,  # what values do you want to change?
                       new,  # what should the new values be?
                       otherNA = TRUE,   # should other values be converted to NA?
                       numeric = TRUE) { # should result be numeric?
   x.new <- x  # copy vector to x.new
   if(class(x.new) == "factor") {x.new <- paste(x.new)} # remove factors
   for(i in 1:length(old)) { # loop through all old values:
    x.new[x == old[i]] <- new[i]
    }
   if(otherNA) { # convert unspecified values to NA:
    x.new[(x %in% old) == FALSE] <- NA
    }
   if(numeric) {x.new <- as.numeric(x.new)}  # convert vector to numeric values
   return(x.new)  # return new vector
}

# Create appropriate subset of 'iris'
iris_anova_prep1 <- iris[,1:4]
iris_anova_prep1$id <- c(1:150)

# Rename and recode 'TCategory' for better readability
iris_anova_prep1$Species <- recode.var(iris$TCategory,
                                       old = c("I.setosa", 
                                               "I.versicolor", 
                                               "I.virginica"),
                                       new = c("Setosa", 
                                               "Versicolor", 
                                               "Virginica"), 
                                       otherNA = FALSE,
                                       numeric = FALSE)
# Save 'Species' as factor variable
iris_anova_prep1$Species <- as.factor(iris_anova_prep1$Species)

# Reformat dataframe from wide to long format with temporary variable 'wtifl_prep'
iris_anova_prep1 <- gather(iris_anova_prep1, 
                           wtifl_prep, 
                           Measurement, 
                           Sepal_length:Petal_width, 
                           factor_key=TRUE)
head(iris_anova_prep1)

# Prepare creation of within-factor variables 'Part_of_Flower' and 'Kind_of_Measurement'
iris_anova_prep2 <- iris_anova_prep1
iris_anova_prep2$Part_of_Flower <- iris_anova_prep2$wtifl_prep
iris_anova_prep2$Kind_of_Measurement <- iris_anova_prep2$wtifl_prep

# Split temporary variable 'wtifl_prep' into within-factor variables 'Part_of_Flower' and 'Kind_of_Measurement'
iris_anova_prep2$Part_of_Flower <- 
  recode.var(iris_anova_prep2$Part_of_Flower,
             old = c("Sepal_length", 
                     "Sepal_width", 
                     "Petal_length", 
                     "Petal_width"),
             new = c("Sepal", 
                     "Sepal", 
                     "Petal", 
                     "Petal"),
             otherNA = FALSE,
             numeric = FALSE)
iris_anova_prep2$Kind_of_Measurement <- 
  recode.var(iris_anova_prep2$Kind_of_Measurement,
             old = c("Sepal_length", 
                     "Sepal_width", 
                     "Petal_length", 
                     "Petal_width"),
             new = c("Length", 
                     "Width", 
                     "Length", 
                     "Width"),
             otherNA = FALSE,
             numeric = FALSE)

# Save 'Part_of_Flower' and 'Kind_of_Measurement' as factor variables
iris_anova_prep2$Part_of_Flower <- as.factor(iris_anova_prep2$Part_of_Flower)
iris_anova_prep2$Kind_of_Measurement <- as.factor(iris_anova_prep2$Kind_of_Measurement)

# Delete temporary variable 'wtifl_prep'
iris_anova <- iris_anova_prep2[,c(1:2,5,6,4)]
head(iris_anova)

```

### Creating the ANOVA Model {#creating-the-anova-model}

```{r}
#| label: anova-model
#| tidy: true
#| tidy-opts:
#|   - width.cutoff=60
#| warning: false
#| message: false

iris_anova_model <- aov_ez("id", 
                           "Measurement", 
                           iris_anova, 
                           between = "Species", 
                           within = c("Part_of_Flower", "Kind_of_Measurement"))

```

*Note. Type III Sums of Squares are used for the calculation of the ANOVA*

### Testing Assumptions of the ANOVA *R* {#testing-assumptions-anova-r}

#### Observations are Independent and Identically Distributed {#oiid}

This assumption cannot be tested empirically, but needs to hold on conceptual or logical grounds. For the purpose of this exercise I am therefore simply assuming, that said assumption holds.

#### Homogeneity of Variances {#homogeneity-of-variances}

The variances across all the groups (cells) of between-subject effects should be the same. In order to test this assumption **Levene's Test** is conducted.

```{r}
#| label: anova-assumtion-test1
#| tidy: true
#| tidy-opts:
#|   - width.cutoff=60
#| warning: false
#| message: false

check_homogeneity(iris_anova_model)

```

These results indicate that the assumption of homogeneity of variances is in fact **significantly violated**.

#### Sphericity {#sphericity}

For within-subjects effects, sphericity is the condition where the variances of the differences between all possible pairs of within-subject conditions (i.e., levels of the independent variable) are equal. In order to test this assumption **Mauchly's test** is conducted.

```{r}
#| label: anova-assumtion-test2
#| tidy: true
#| tidy-opts:
#|   - width.cutoff=60
#| warning: false
#| message: false

check_sphericity(iris_anova_model)

```

These results indicate that the assumption of sphericity is **not significantly violated**.

#### Normality of Residuals {#normality-of-residuals}

The errors used for the estimation of the error term(s) (MSE) are normally distributed. In order to test this assumption the **Shapiro-Wilk test** is conducted.

```{r}
#| label: anova-assumtion-test3
#| tidy: true
#| tidy-opts:
#|   - width.cutoff=60
#| warning: false
#| message: false

iris_anova_norm <- check_normality(iris_anova_model)
iris_anova_norm
```

Since this test tends to have high type-I error rates, a visual inspection of the residuals using a **quantile-quantile plot** (qq-plots) is preferred.

```{r}
#| label: anova-assumtion-test4
#| tidy: true
#| tidy-opts:
#|   - width.cutoff=60
#| warning: false
#| message: false

plot(iris_anova_norm, type = "qq")

```

Figure 2.1: Quantile-Quantile plot

If the residuals were normally distributed, we would see them falling close to the diagonal line, inside the 95% confidence bands around the qq-line. Unfortunately it seems, that the assumption of the normality of residuals is in deed **significantly violated**.

#### Conclusion {#conclusion-assumptions}

Since this is just an exercise in conducting ANOVA analyses, the above results are simply ignored and the analysis was conducted nonetheless.

### Results of Mixed ANOVA {#results-mixed-anova}

```{r}
#| label: anova-analysis
#| tidy: true
#| tidy-opts:
#|   - width.cutoff=60
#| warning: false
#| message: false

summary(iris_anova_model)
```
